Formal Verification Tool Based on Symbolic Execution for Smart Contract

### 2023-11-28 papers/static-checker/ecoop2024/ecoop2024.tex
add instructions parts

\begin{mathpar}
\inferrule[]
  { \NEG\ \PREDICATE
  }{
  \{[\INSTRUCTION, \STACK, \PREDICATE]\} \cup \SYSTEM \SystemTrans \SYSTEM}
\end{mathpar}
There are various types of instructions in Michelson, categorized based on their functions. These include control structures, stack manipulation, arithmetic operations, boolean operations, cryptographic operations, and operations on data structures. In our model, we classify them into three main groups: basic instructions, branch instructions, and loop instructions. This categorization reflects how these instructions are handled within our specific model.
\subsubsection{Basic Instructions}

@@

@@

%FAILWITH
\begin{mathpar}
  \inferrule[FAILWITH]
  {
  }{[(\FAILWITH; \INSTRUCTION), (\StackOne,  \TY) \STACKCONCAT \STACK,  \PREDICATE] \StateTrans [\EMPTYSTACK, failwith (\StackOne) \STACKCONCAT\EMPTYSTACK, \PREDICATE]}
\end{mathpar}
\subsection{Dealing with loops}
\subsubsection{For loop: ITER}
\subsubsection{While loops: \LOOP\ and \LOOP-LEFT}
\section {Domain Specific Language}
\section {Static Checker}
\section {Case Studies}
\subsection{USDtz}
\subsection{Kolibri Oracle Contract}
\section {Related Work}
\section {Conclusion}
\section {Appendix}

>>>>

\begin{mathpar}
\inferrule[]
  { \NEG\ \PREDICATE
  }{
  \{[\INSTRUCTION, \STACK, \PREDICATE]\} \cup \SYSTEM \SystemTrans \SYSTEM}
\end{mathpar}

There are various types of instructions in Michelson, categorized based on their functions. These include control structures, stack manipulation, arithmetic operations, boolean operations, cryptographic operations, and operations on data structures. In our model, we classify them into three main groups: basic instructions, branch instructions, and loop instructions. This categorization reflects how these instructions are handled within our specific model.
\subsubsection{One-step Instructions}
A one-step instruction only effects a system state itself, does not produce a new state, and halts the execution after a single operation. The ABS rule specifies the corresponding instruction that applies to the first element of the stack. This element must have the type of int and returns the positive value of its current value, which is of a natural number type. Another example, the COMPARE instruction compares the two first elements of a stack, both of which have a natural number type. It returns a result of either -1, 0, or 1 depending on the comperations of their values.
%ABS
\begin{mathpar}
\inferrule[ABS]
  {
  }
  {
    [(\ABS; \INSTRUCTION), (\StackOne, \TINT) \STACKCONCAT \STACK,
    \PREDICATE]
    \StateTrans \
    [\INSTRUCTION, (\VariableX, \TNAT) \STACKCONCAT \STACK,
    \PREDICATE \wedge (\StackOne \ge 0 \Rightarrow \VariableX =
    \StackOne) \wedge (\StackOne <0 \Rightarrow \VariableX = -\StackOne)]
 }
\end{mathpar}

% COMPARE
\begin{mathpar}
\inferrule[COMPARE]
  {
  }
  {
    [(\COMPARE ; \INSTRUCTION), (\StackOne, \TNAT) \STACKCONCAT (\StackTwo, \TNAT)
    \STACKCONCAT \STACK, \PREDICATE ]
    \SystemTrans \\
    [\INSTRUCTION, (\VariableX, \TINT) \STACKCONCAT \STACK, \PREDICATE
    \wedge\ (\StackOne > \StackTwo \Leftrightarrow \VariableX = 1)
    \wedge\ (\StackOne = \StackTwo \Leftrightarrow \VariableX = 0) 
    \wedge\ (\StackOne < \StackTwo \Leftrightarrow \VariableX = -1)]
    }
\end{mathpar}
\subsubsection{Multi-step Instructions}
Multi-step instructions require the execution of a sub-sequence of instructions before returning to the main execution. The following rule models the instruction EXEC, which executes a function specified as a sequence of instructions denoted as \INSTRUCTIONONE\ in the rule. This instruction first applies the code of the function to the first element of the stack. It then places the result back into the main execution.
\begin{mathpar}
  \inferrule[EXEC]
  {
    [\INSTRUCTIONONE, (\StackOne, \TYF) \STACKCONCAT \EMPTYSTACK, 
    Q]
    \StateTrans^*
    [ \EMPTYSTACK, (\StackOne', \TYS) \STACKCONCAT Q']
  }{
     [(\EXEC; \INSTRUCTION),   (\{\INSTRUCTIONONE\}, \TYF\ \rightarrow\ \TYS) \STACKCONCAT
    (\StackOne, \TYF) \STACKCONCAT \STACK, \PREDICATE \wedge Q] 
    \StateTrans \
    [ \INSTRUCTION, (\StackOne', \TYS) \STACKCONCAT \STACK,
    \PREDICATE \wedge Q']}
\end{mathpar}

The \DIP\ \N\ instruction delves into the first \N\ elements of a stack, applies the sequence of instructions denoted as \INSTRUCTIONONE\ to the remaining part of the stack, modifies a part of the predicate and subsequently pushes the result back into the main execution.
%DIP n
\begin{mathpar}
\inferrule[\DIP\ \N]
  { 
     \FLEN(\A) \EQUAL\ \N \\ [\INSTRUCTIONONE,  \B, Q]
    \StateTrans^*
    [\EMPTYSTACK,  \B_1, Q']
  }
  {[(\DIP\ \N\ \INSTRUCTIONONE; \INSTRUCTION), \A\ \At\ \B, \PREDICATE \wedge Q] \StateTrans 
[(\INSTRUCTION), \A\ \At\ \B_1, \PREDICATE \wedge Q']}
\end{mathpar}
\subsubsection{Blockchain Instructions}
 Blockchain instructions make implicit use of data from an execution environment, derived from the current state of the blockchain (the context), from the transaction that triggered the contract, and from the block containing this transaction. We model these data as symbolic values, which satisfis some constrants. For example, the AMOUNT instruction adds to the stack the amount of tokens that sent together with the transcation which calls this smart contract. Here, the amount is modeled as a symbolic value donated as \VAMOUNT, which has \TMUTEZ\ type, which is a positive number value.

\begin{mathpar}
\inferrule[AMOUNT]
  {
  }
  {[(\AMOUNT; \INSTRUCTION), \STACK, \PREDICATE] \StateTrans \
[\INSTRUCTION, (\VAMOUNT, \TMUTEZ) \STACKCONCAT\STACK, \PREDICATE}
\end{mathpar}

\subsubsection{Cryptographic Instructions}
Cryptographic instructions in Michelson utilize built-in functions that manage cryptographic features. These functions are symbolically modeled in our framework, specifying their inputs and outputs. For instance, the instruction \SHA\ computes the cryptographic hash of the top of the stack using the SHA-256 cryptographic hash function. The function SHA-256 is symbolically represented as the function  \FSHA\  that takes an input of type \TBYTE\ and returns another byte value, constrained to a length of 256.
%HASH-KEY
\begin{mathpar}
\inferrule[\SHA]
  {
  }
  {[(\SHA; \INSTRUCTION), (\StackOne, \TBYTE) \STACKCONCAT\STACK, \PREDICATE] \StateTrans \
[\INSTRUCTION, (\VariableX, \TBYTE) \STACKCONCAT\STACK, \PREDICATE \Wedge\ (\VariableX\ = \FSHA(\StackOne))]}
\end{mathpar}
\subsubsection{Branch Instructions}
The Michelson language incorporates conditional branching, enabling the execution of specific code segments based on certain criteria. One of the instructions facilitating this is the IF {} {} construct. For example, the \IF\ \INSTRUCTIONONE\  \INSTRUCTIONTWO\ instruction allows the creation of branches in execution. It takes two sequences as arguments. Expecting a boolean value at the top of the stack, it consumes this top element. If the boolean is True, it executes the first provided sequence; otherwise, it executes the second sequence.
%IF
\begin{mathpar}
  \inferrule[IF]
  {  
  }{
    \{[(\IF\ \INSTRUCTIONONE\  \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TBOOL) \STACKCONCAT\STACK, \PREDICATE]\} \cup \SYSTEM 
      \SystemTrans 
    \{[\INSTRUCTIONONE, \STACK, \PREDICATE\ \Wedge\ \StackOne]\} \cup \{ [\INSTRUCTIONTWO, \STACK, \PREDICATE\ \Wedge\ \NEG\
   \StackOne]\} \cup  \SYSTEM
  }
\end{mathpar}

The \IFLEFT\ \INSTRUCTIONONE\  \INSTRUCTIONTWO\ instruction expects a or value at the top of the stack, it consumes this top element. If the value is Left, it executes the first provided sequence; otherwise, it executes the second sequence.
%IF-LEFT
\begin{mathpar}
  \inferrule[IF-LEFT]
  {  
  }{
    \{[(\IFLEFT\ \INSTRUCTIONONE\ \INSTRUCTIONTWO; \INSTRUCTION),
    (\StackOne, \TOR\ \TYF\ \TYS) \STACKCONCAT \STACK, \PREDICATE]\}  \cup  \SYSTEM
    \SystemTrans \\
    \{[\INSTRUCTIONONE, (\VariableX, \TYF) \STACKCONCAT\STACK,
    \PREDICATE \wedge (\StackOne\ \EQUAL\ \LEFT\ \VariableX)]\}  \cup  \{  [\INSTRUCTIONTWO, (\VariableX, \TYS) \STACKCONCAT\STACK, \PREDICATE \wedge (\StackOne\ \EQUAL\ \RIGHT\ \VariableX))]\} \cup \SYSTEM
  }
\end{mathpar}

\subsubsection{Loop Instructions}
In Michelson, there are several loop instructions that execute a loop over the data structure, such as ITER, MAP, MEMBER, SIZE and UPDATE on list, set, map or on the stack, such as LOOP and LOOP-LEFT.

%%ITER
\begin{mathpar}
  \inferrule[ITER-empty]
  {  
  }{
    [(\ITER\ \INSTRUCTIONONE ; \INSTRUCTION), (\StackOne, \TYLIST\ \TY) \STACKCONCAT\STACK, \PREDICATE] \StateTrans\ [\INSTRUCTION, \STACK,  \PREDICATE\ \Wedge\  (\StackOne\ \EQUAL\ \EMPTYLIST)]
  }
\end{mathpar}

@@

Let us consider the \ITER\ instruction (for list), which applies to a stack \STACK, where the top element has type  (\TY\ \TYLIST) and the rest of the stack has type \TYA, returns the result stack of type \TYA. The \ITER\ instruction loops on the datastructure list. 
Typing
\begin{mathpar}
  \inferrule{\JTypeExpr\TEnv{\INSTRUCTION}{\TY : \TYA\ \SRightarrow\ \TYA}
  }{
      \JTypeExpr\TEnv{\ITER\ \INSTRUCTION}{\TYLIST\ \TY : \TYA\ \SRightarrow\ \TYA}
    }
\end{mathpar}

Semantic

\begin{mathpar}
  \inferrule{
  }{
    \ITER\ \INSTRUCTION\ / \{ \} \STACKCONCAT\ \STACK\ \SRightarrow\  \STACK
    }
\end{mathpar}

\begin{mathpar} 
  \inferrule{\INSTRUCTION\  /  \VariableX\ \STACKCONCAT\ \STACK \SRightarrow\ \STACK'
  }{
    \ITER\ \INSTRUCTION\ / \{ \VariableX\ ; \TAIL\ \} \STACKCONCAT\ \STACK\ \SRightarrow\ \ITER\ \INSTRUCTION\ / \{ \TAIL\ \} \STACKCONCAT\ \STACK' 
    }
\end{mathpar}

If a list \LIST\ is concrete, the intructuion simply loops until the list ends with an empty list construction. Here we consider the synbolic execution of \ITER\ when the list is a sybolic term or variable.

The sequence of intructuions I is a lamda function  (\TLAMBDA\ (\TY\ : \TYA)\ \TYA) applied to the stack of type (\TY\ : \TYA), returning the stack of type \TYA. We can represent I as the function f that takes an element of type \TY\ and a stack of type \TYA\ and returns a stack of type \TYA.

\begin{mathpar}
  f : \TY\  \TYA\ \SRightarrow\ \TYA
\end{mathpar}

\begin{mathpar}
  f  \VariableX\ \STACK =  (\TLAMBDA\ (\TY\ : \TYA)\ \TYA) (\VariableX\  \STACKCONCAT\ \STACK)
\end{mathpar}

or

\begin{mathpar}
  f  \VariableX\ \STACK = I / (\VariableX\  \STACKCONCAT\ \STACK)
\end{mathpar}

Let  
\LIST\ \STACKCONCAT\ \STACK

 be the stack just before applying the instruction ITER I, where \VariableX\ be the head of the list and \STACK\ be the rest of the stack . For first  loop, let  \STACK\ be f (\VariableX, \STACK). \STACK\ is the stack resulting from applying the function f to the element \VariableX\ and the stack \STACK. In other words, the stack \STACK\ is formed by applying the sequence of the instructions I to the stack \VariableX\ \STACKCONCAT\ \STACK.

Let the result of applying the instruction \ITER\ I on the stack \LIST\ ::  \STACK\ be the stack S', then Z is resulted from the \FOLD\ function that applies the function f  to each element of the list \LIST\ and initial value is the stack \STACK.

\begin{mathpar}
S' =  \FOLD\ \STACK\ f   \LIST
\end{mathpar}

@@

%LOOP
\begin{mathpar}
  \inferrule[LOOP-true]
  {  
  }{
    [(\LOOP\ \INSTRUCTIONONE; \INSTRUCTION),  (\StackOne, \TBOOL)
    \STACKCONCAT\STACK, \PREDICATE]
    \StateTrans\
    [(\INSTRUCTIONONE; \LOOP\ \INSTRUCTIONONE; \INSTRUCTION),
    \STACK, \PREDICATE \wedge \StackOne]
  }

  \inferrule[LOOP-false]
  {  
  }{
    [(\LOOP\ \INSTRUCTIONONE; \INSTRUCTION),  (\StackOne, \TBOOL) \STACKCONCAT\
    \STACK, \PREDICATE]
    \StateTrans\
   [\INSTRUCTION, \STACK, \PREDICATE \wedge
   (\NEG\StackOne)]
   }
\end{mathpar}

\subsubsection{FAILWITH}
a Michelson program can fail, explicitly using a specific opcode

%FAILWITH
\begin{mathpar}
  \inferrule[FAILWITH]
  {
  }{[(\FAILWITH; \INSTRUCTION), (\StackOne,  \TY) \STACKCONCAT \STACK,  \PREDICATE] \StateTrans [\EMPTYSTACK, failwith (\StackOne) \STACKCONCAT\EMPTYSTACK, \PREDICATE]}
\end{mathpar}
\section {Domain Specific Language}
\section {Static Checker}
\section {Case Studies}
\subsection{USDtz}
\subsection{Kolibri Oracle Contract}
\section {Related Work}
\section {Conclusion}
\section {Appendix}
